%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% chapter1.tex
%% NOVA thesis document file
%%
%% Chapter with introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE chapter1.tex}%

\chapter{Introduction}
\label{cha:introduction}

\prependtographicspath{{\DIRFIGURES/Covers/}}

% epigraph configuration
\epigraphfontsize{\small\itshape}
\setlength\epigraphwidth{12.5cm}
\setlength\epigraphrule{0pt}


\section{Context and motivation}
\label{sec:a_bit_of_history}

\ntindex[Template]{}

The global imperative to mitigate climate change has placed the transportation sector at the center of sustainability debates, as it remains one of the largest contributors to greenhouse gas (GHG) emissions worldwide, it accounts for roughly 23\%\ of total greenhouse gas emissions \parencite{IPCC2022Transport}. Within this sector, Urban Freight Transportation (UFT) plays a critical yet often underestimated role. Urban freight activities are essential to the functioning of modern cities, ensuring that goods are efficiently delivered to businesses and consumers. At the same time, they significantly contribute to traffic congestion, local air pollution, and carbon emissions, particularly in densely populated urban areas \parencite{HAO201594}. Consequently, urban freight systems are increasingly subject to international decarbonization targets and stringent local environmental regulations.

In response, public authorities across the world have adopted a wide range of policy measures to manage and decarbonize urban freight transport. A comprehensive international review by \textcite{HolguinVeras2015a, HolguinVeras2015b}, shows that cities across Europe, North America, Asia, and Latin America have implemented initiatives such as access restrictions, low-emission zones, vehicle-related regulations, traffic management schemes, and infrastructure adaptations. Evidence from 56 cities in 32 countries indicates that regulatory and supply-side approaches dominate current practice, with Europe and Asia leading the implementation of vehicle-related strategies, such as emission standards and low-emission zones. In contrast in North America relies more heavily on infrastructure and operational management measures rather than vehicle-restriction policies, such as Last Mile Delivery Practices, Demand Management, Intelligent Transportation Systems (ITS).

Although these policies have been effective in reducing emissions, congestion, and other negative externalities, they frequently impose additional costs on freight operators. In particular, as mentioned by \textcite{BroaddusBrowneAllen2015} and \textcite{Watkiss2003LondonLEZ} vehicle-related restrictions and low-emission zones are associated with increased delivery costs and accelerated fleet renewal, highlighting a fundamental trade-off between environmental objectives and operational profitability. As a result, \textcite{Ahani2023UFT}, shows that urban freight policies increasingly transfer financial and investment risk to private fleet operators, who must adapt their vehicle portfolios under regulatory pressure while facing uncertain economic returns and to address this problem, policymakers should be responsible to foresee policies which would encourage and support UFT companies during fleet replacement.

Despite the breadth and intensity of these policy interventions, their translation into large-scale fleet transformation has been uneven. A systematic literature review by \textcite{KervallPalsson2022}, covering 93 peer-reviewed studies, identifies a wide range of barriers that hinder the transformation of urban freight systems, including economic constraints, technological uncertainty, lack of charging infrastructure, and policy misalignment. Among these, economic barriers and the so-called “first-mover disadvantage” are particularly critical for fleet operators, as early adopters of electric vehicles face high upfront investment costs, uncertain operational benefits, and volatile energy prices. As a consequence, fleet replacement decisions are often postponed or limited to incremental changes, despite strong regulatory pressure.

This combination of regulatory ambition and systemic barriers transforms fleet replacement into a complex decision-making problem under uncertainty. While electric vehicles offer clear environmental benefits and the potential for lower operating costs, they are associated with high capital expenditures and significant technological uncertainty. Fleet managers must make high-stakes investment decisions in a volatile environment, where fuel prices, electricity costs, and vehicle residual values fluctuate unpredictably (\parencite{IEA2026OilMarketReport,Brockmann2025ElectricityVolatility,Gautam2024EVDepreciation}). The central challenge, therefore, is not merely environmental compliance, but determining an optimal fleet composition that balances conventional and alternative vehicle technologies while meeting sustainability targets without compromising financial viability.

\section{Research Questions}
\label{sec:Research_Questions}

To address the identified gaps and guide the development of the proposed optimization framework, this thesis aims to answer the following core research questions:

\begin{itemize}
    \item \textbf{RQ1:} How does the trade-off between Expected Total Cost and cost Variance influence the optimal composition of an urban freight fleet under uncertainty?
    
    \item \textbf{RQ2:} How do different levels of risk aversion, represented through a mean–variance framework, affect fleet replacement decisions and the resulting efficient frontier?
    
    \item \textbf{RQ3:} How does the trade-off between Expected Total Cost and cost Variance,
    as represented by the efficient frontier, influence the optimal transition from conventional
    vehicle technologies to electric and hybrid alternatives in urban freight fleets?
    
    \item \textbf{RQ4:} How sensitive are the Pareto-optimal fleet configurations to variations
    in key market and policy parameters (energy prices, vehicle acquisition costs, and
    carbon pricing), and which parameter thresholds trigger significant shifts in optimal fleet
    composition?
\end{itemize}

\section{Previous approach and objectives}
\label{sec:Previous_approach_and_objectives}

To support investment decisions under uncertainty in fleet replacement problems, the financial literature offers well-established quantitative tools, most notably Modern Portfolio Theory (MPT), originally introduced by \textcite{Markowitz1952}. MPT provides a framework for balancing expected performance and risk by treating different assets as components of a portfolio with distinct cost and risk characteristics. This framework has been successfully adapted to the context of urban freight transportation by \textcite{Ahani2016}, who modeled vehicle technologies as portfolio assets and demonstrated that a diversified fleet composition can hedge against uncertainty in fuel prices and operating costs. In the formulation proposed, the fleet replacement problem is modeled as a multi-objective optimization problem, with the objectives of minimizing the Expected Total Cost (ETC) and minimizing the Variance of the total cost. The expected cost captures average economic performance, while variance is used as a measure of risk, reflecting cost volatility arising from uncertain market conditions. This mean–variance formulation allows decision-makers to explicitly analyze the trade-off between cost efficiency and risk exposure, providing a structured basis for fleet composition decisions under uncertainty.

Despite its relevance and practical applicability, existing risk-aware applications in fleet replacement including mean-variance models, often rely on simplified representations of real-world fleets (\parencite{Pelletier2019EBFTP, Malladi2020}). In particular, many models focus on a limited number of vehicle technologies or assume homogeneous vehicle characteristics, which restricts their ability to capture the diversity observed in actual urban freight fleets. Moreover, the increasing availability of alternative vehicle technologies, such as electric, diesel, and hybrid vehicles with different sizes and operational profiles, calls for more comprehensive modeling approaches that can reflect this heterogeneity while preserving analytical tractability.

Building on the framework of \textcite{Ahani2016}, this work extends the mean–variance approach by developing a more detailed fleet replacement optimization model that accommodates multiple vehicle technologies and characteristics. The resulting problem is formulated as a bi-objective mixed-integer quadratic programming (MIQP) model. To solve this problem, the \texorpdfstring{$\varepsilon$}{ε}-constraint method is employed, whereby one objective is optimized while the other is constrained within predefined bounds. By systematically varying the $\varepsilon$ parameter, the efficient frontier of optimal fleet compositions can be generated, enabling a clear visualization and analysis of the trade-offs between expected cost and risk.

Although variance is a convenient and widely used proxy for risk, it captures overall dispersion and may fail to distinguish between moderate variability and extreme downside outcomes. For this reason, the portfolio optimization literature proposes several alternative risk measures with different mathematical properties and computational implications. Linear deviation-based measures such as Mean Absolute Deviation (MAD) can approximate risk while preserving linear structure, often leading to MILP formulations that scale better than quadratic variance models \parencite{MANSINI2014518}. Tail-focused measures such as Conditional Value-at-Risk (CVaR) explicitly target extreme losses and are widely adopted in scenario-based optimization because they can be formulated using linear constraints \parencite{RockafellarUryasev2000}. More recently, non-linear coherent measures such as Entropic Value-at-Risk (EVaR) have been proposed to overcome limitations of VaR/CVaR monotonicity and to provide tractable convex formulations in sample-based settings \parencite{AhmadiJavid2011}. Collectively, these measures highlight that "risk" can be operationalized in multiple linear and non-linear ways, motivating the broader risk perspective adopted in this dissertation \parencite{PichlerSchiller2015}.

The primary objective of this thesis is to develop and analyze a risk-aware optimization framework for urban fleet replacement decisions under uncertainty, using mean-variance as the baseline formulation and positioning it relative to alternative linear and non-linear risk measures. Specifically, this work aims to:

\begin{itemize}
    \item Formulate a multi-objective optimization model that minimizes Expected Total Cost and a quantitative risk measure (baseline: cost Variance) in urban fleet replacement decisions;
    \item Extend existing models by incorporating multiple vehicle technologies and fleet characteristics;
    \item Apply the $\varepsilon$-constraint method to solve a sequence of mixed-integer quadratically constrained problems (MIQCPs) and generate the efficient frontier;
    \item Analyze the trade-offs between cost and risk under different fleet compositions and market conditions;
    \item Contextualize the baseline variance risk model by discussing alternative linear (e.g., MAD) and non-linear/tail-oriented risk measures (CVaR/EVaR) and their tractability implications for fleet planning.
\end{itemize}

In this context, the proposed framework is intended to support both private and public decision-makers by providing a transparent and quantitative tool for evaluating fleet replacement strategies that balance economic performance and risk in the transition toward more sustainable urban freight systems.

%% ---------------------------- -------------------------------- -----------------------------
%% ---------------------------- -------Document Structure------- -----------------------------
%% ---------------------------- -------------------------------- -----------------------------
\section{Document Structure}
\label{sec:Document_Structure}

The remainder of this dissertation is organized as follows.

Chapter 2 presents a comprehensive literature review on sustainable urban freight transportation, fleet replacement optimization models, and the application of portfolio theory to transportation systems. Special attention is given to mean–variance formulations and risk modeling approaches relevant to fleet investment decisions under uncertainty.

Chapter 3 describes the case study considered in this work. It introduces the characteristics of the urban freight fleet, the operational context, and the data assumptions used to represent vehicle technologies, cost components, and uncertainty sources.

Chapter 4 presents the proposed optimization framework. This chapter details the mathematical formulation of the multi-objective fleet replacement model, including decision variables, objective functions, and constraints. The solution approach based on the $\varepsilon$-constraint method is also described, along with the procedure used to generate the efficient frontier.

Chapter 5 outlines the tasks to be carried out for the remainder of this dissertation, along with the planned schedule for their completion.

\newpage